---
title: "Importing and exporting data"
author: "UC Davis Bioinformatics Core"
date: "10/21/2021"
output: html_document
---

# Importing and exporting data

Manual data entry is time-consuming and leads to errors. R has a number of functions for reading data in a variety of formats. Let's use the `read.csv()` function to read in a spreadsheet containing data from an experiment.

```{r}
birthweight <- read.csv("datasets/birthweight.csv")
```

CSV stands for "comma separated value," and the CSV file is simply a text file where each row in the file represents a row in the data table, and the columns are separated by commas. The contents of the CSV file are now stored in the variable "birthweight."

If we wanted to write the contents of the birthweight object to a new CSV, we can use the `write.csv()` function instead.

```{r}
write.csv(birthweight, file = "datasets/new_birthweight.csv")
```

The similar `read.delim()` and `write.delim()` can be used to read and write tab-delimited files, where columns are separated by tab characters rather than commas.

